---
title: 'Analysis: Change in weight over 3 months'
author: "Julianna Renzi"
date: "2026-03-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r clear environment}
# clear environment for reproducibility
rm(list=ls())
```

```{r load packages}
# load librarian (package for installing/loading packages)
if (!require("librarian")) install.packages("librarian")

# load required packages
librarian::shelf(here, # relative file paths 
                 tidyverse, # data wrangling
                 glmmTMB, # model creation
                 car, # type II ANOVA
                 emmeans, # posthoc contrasts
                 performance, # model assumptions
                 multcomp, # cld
                 ggtext # for complicated axes labels
                 )
```

```{r set color palette}
"#E1BE6A" -> no.snail.col
"#40B0A6" -> snail.col

"#c49627" -> no.snail.col.dark
"#2d7d76" -> snail.col.dark
```

# Read in data

```{r}
# coral wet weight data over the macroalgal addition time period
weights_1 <- read_csv(here("data", "coral_wet_weights_3mo.csv"))

# metadata for corals 
metadata <- read_csv(here("data", "metadata.csv"))
```

# Calculate weight change

```{r}
weights_1 %>% 
  pivot_wider(id_cols = coral_id,
              names_from = time_point,
              values_from = wet_weight_g) %>% 
  # calculate percent change
  mutate(percent_change = ((`1`-`0`)/`0`) * 100) %>% 
  # join with metadata
  full_join(metadata) %>% 
  # make structure labels nice for plotting
  mutate(treatment_skeleton = case_when(treatment_skeleton == "control" ~ "Control",
                                        treatment_skeleton == "hali" ~ "Halimeda skeleton",
                                        treatment_skeleton == "loboph" ~ "Lobophora skeleton",
                                        treatment_skeleton == "skeleton" ~ "Skeleton")) %>% 
  # get into a nice order
  mutate(treatment_skeleton = factor(treatment_skeleton, levels = c("Control", "Skeleton", "Halimeda skeleton", "Lobophora skeleton"))) -> weights_2
```

# Model weight change

## With an interaction

```{r}
mW1 <- glmmTMB(percent_change ~ treatment_skeleton + treatment_drupella + 
                 treatment_skeleton:treatment_drupella + `0` + (1|parent_colony),  
              data = weights_2)
```

Examine outputs and check whether interaction was important

```{r}
summary(mW1)
car::Anova(mW1) # interaction is not important 
```

## Without the interaction

Drop the interaction because it does not appear important 

```{r}
mW2 <- glmmTMB(percent_change ~ treatment_skeleton + treatment_drupella + `0` + (1|parent_colony), 
              data = weights_2)
```

Examine outputs and check whether interaction was important

```{r}
summary(mW2)
car::Anova(mW2) # interaction is not important 
```

Check model 

```{r}
check_model(mW2) # looks good
```

Examine posthoc contrasts

```{r}
# plot marginal means for skeleton treatment to see expected ordering
plot(emmeans(mW2, ~ treatment_skeleton ))
```

And magnitudes

```{r}
# examine posthoc contrasts (pairwise is a shortcut for pairs())
emmeans(mW2, pairwise ~ treatment_skeleton, 
        type = "response") # this is overkill (no link function and Gaussian)

emmeans(mW2, pairwise ~ treatment_drupella, 
        type = "response") # this is overkill (no link function and Gaussian)
```

# Plot model and data

Get model dataframe. First, get compact letter displays from the package `cld` and pairwise `emmeeans` comparisons

```{r}
grps <- as_tibble(cld(emmeans(mW2, pairwise ~ treatment_skeleton | treatment_drupella, 
                              type = "response"), 
                      Letters = letters)) %>% 
  # make a nicer grouping variable without spaces
  mutate(.group = str_remove_all(.group, " ")) %>% 
  rename(group = .group) %>% 
  # add spaces for plotting aesthetics
  mutate(group = case_when(group == "a" ~ "  a",
                           group == "c" ~ "  c",
                           TRUE ~ group))
```

Then join with the estimated marginal means and errors from the main `emmeans` call

```{r}
as_tibble(emmeans(mW2, pairwise ~ treatment_skeleton | treatment_drupella, 
                  # response is overkill here as well 
                  type = "response")$emmeans) %>% 
  full_join(grps) -> grps_2
```

Plot together with raw data and save as a PDF in our `figures` folder

```{r}
pdf(here("figures", "fig_2_weights.pdf"), height = 3, width = 5.25)

grps_2 %>% 
  ggplot() +
  geom_pointrange(aes(x = treatment_skeleton, y = emmean, 
                      color = treatment_drupella,
                      ymin = lower.CL, ymax = upper.CL),
                      size = .8, linewidth = .8,
                  position = position_dodge(width = 0.5)) +
  scale_color_manual(name = "Snail wound", labels = c("n" = "No", "y" = "Yes"),
                     values = c(no.snail.col, snail.col)) +
  # make a new color scale
  ggnewscale::new_scale_color() +
  # add raw data
  geom_jitter(data = weights_2,
              aes(x = treatment_skeleton, 
                  y = percent_change, 
                  # need to use fill here with shape 21 to force the offset 
                  fill = treatment_drupella,
                  stroke = .5, # controls size of outline
                  fill = NA, # don't actually want the fill so we can see the points well (just for grouping)
                  color = treatment_drupella),
              size = 2, shape = 21,
              position = position_jitterdodge(
                jitter.width = 0.15,
                jitter.height = 0,
                dodge.width = 0.5 # matching the position_dodge() above
              )
  ) +
  scale_color_manual(name = "Snail wound", labels = c("n" = "No", "y" = "Yes"),
                     values = c(no.snail.col.dark, snail.col.dark)) +
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "top", axis.text.x = element_markdown(lineheight = 1)) +
  ylab("Weight change (%)") +
  xlab("") +
  scale_x_discrete(
    labels = (c("Control" = "Control", "Skeleton" = "Skeleton",
                "Halimeda skeleton" = "<i>Halimeda</i><br>skeleton",
                "Lobophora skeleton" = "<i>Lobophora</i><br>skeleton"))) +
  ggnewscale::new_scale_color() +
  geom_text(data = subset(grps_2, treatment_drupella == "n"),
            aes(label = group, x = treatment_skeleton, y = emmean), 
              vjust = -1.5, hjust = 3, size = 4, color = "black") 

dev.off()
```



