---
title: 'Analysis: Change in weight over 3 months'
author: "Julianna Renzi"
date: "2026-03-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r clear environment}
# clear environment for reproducibility
rm(list=ls())
```

```{r load packages}
# load librarian (package for installing/loading packages)
if (!require("librarian")) install.packages("librarian")

# load required packages
librarian::shelf(here, # relative file paths 
                 tidyverse, # data wrangling
                 glmmTMB, # model creation
                 car, # type II ANOVA
                 emmeans, # posthoc contrasts
                 performance, # model assumptions
                 multcomp, # cld
                 ggtext, # for complicated axes labels
                 patchwork # for multiple panels
                 )
```

```{r set color palette}
"#E1BE6A" -> no.snail.col
"#40B0A6" -> snail.col

"#c49627" -> no.snail.col.dark
"#2d7d76" -> snail.col.dark
```

# Read in data

```{r}
# coral wet weight data over the macroalgal addition time period
weights_1 <- read_csv(here("data", "coral_wet_weights_3mo.csv"))

# metadata for corals 
metadata <- read_csv(here("data", "metadata.csv"))
```

# Calculate weight change

```{r}
weights_1 %>% 
  pivot_wider(id_cols = coral_id,
              names_from = time_point,
              values_from = wet_weight_g) %>% 
  # calculate percent change
  mutate(change = (`1`-`0`)) %>% 
  # join with metadata
  full_join(metadata) %>% 
  # make structure labels nice for plotting
  mutate(treatment_skeleton = case_when(treatment_skeleton == "control" ~ "Control",
                                        treatment_skeleton == "hali" ~ "Halimeda skeleton",
                                        treatment_skeleton == "loboph" ~ "Lobophora skeleton",
                                        treatment_skeleton == "skeleton" ~ "Skeleton")) %>% 
  # get into a nice order
  mutate(treatment_skeleton = factor(treatment_skeleton, levels = c("Control", "Skeleton", "Halimeda skeleton", "Lobophora skeleton"))) -> weights_2
```

# Model weight change

Also significant/similar conclusions if we use `1` (final weight) or percent weight change instead of actual change 

```{r}
mW1 <- glmmTMB(change ~ treatment_skeleton + treatment_drupella + 
                 treatment_skeleton:treatment_drupella + `0` + (1|parent_colony),  
              data = weights_2)
```

```{r}
check_model(mW1)
```

Examine outputs and which terms were important

```{r}
summary(mW1)
car::Anova(mW1) # interaction is not important 
```

Examine posthoc contrasts visually (for multifactor structure treatment)

```{r}
# plot marginal means for skeleton treatment to see expected ordering
plot(emmeans(mW1, ~ treatment_skeleton))
```

Examine magnitudes (report these in the results)

```{r}
# examine posthoc contrasts (pairwise is a shortcut for pairs())
emmeans(mW1, pairwise ~ treatment_drupella, 
        type = "response") # this is overkill (no link function and Gaussian)

emmeans(mW1, pairwise ~ treatment_skeleton, 
        type = "response") # this is overkill (no link function and Gaussian)
```

# Plot results

Get model dataframe for plotting. First, get pairwise `emmeeans` comparisons and compact letter displays from the package `cld`.

```{r}
emm_skel  <- emmeans(mW1, ~ treatment_skeleton) # marginal means, averaged over snail

emm_snail <- emmeans(mW1, ~ treatment_drupella)   # marginal means, averaged over 
```

Get compact letter displays for structure and snail treatments

```{r}
bind_rows(
  as_tibble(cld(emm_skel, Letters = letters)) %>% 
    rename(level = treatment_skeleton) %>% 
    mutate(factor = "Structure treatment"),
  as_tibble(cld(emm_snail, Letters = letters)) %>% 
    rename(level = treatment_drupella) %>% 
    mutate(factor = "Snail wound")
) %>% 
  # tidy the grouping variable (has those weird spaces)
  mutate(group = str_remove_all(.group, " "),
         level = as.character(level),
         # change labels so italics show on plot
         level_label = case_when(
           level == "Halimeda skeleton"  ~ "<i>Halimeda</i><br>skeleton",
           level == "Lobophora skeleton" ~ "<i>Lobophora</i><br>skeleton",
           level == "n" ~ "No",
           level == "y" ~ "Yes",
           TRUE ~ level),
         level_label = factor(level_label,
                              levels = c("No", "Yes", "Control", "Skeleton", "<i>Halimeda</i><br>skeleton", "<i>Lobophora</i><br>skeleton") )
                              ) %>% 
  dplyr::select(-`.group`) -> marg
```

Make labels for the plot

```{r}
# shared x-axis labels for panel a
skel_labs <- c("Control" = "Control",
               "Skeleton" = "Skeleton",
               "Halimeda skeleton" = "<i>Halimeda</i><br>skeleton",
               "Lobophora skeleton" = "<i>Lobophora</i><br>skeleton")
```

Make panel a (raw data in boxplot form):

```{r}
weights_2 %>% 
  ggplot() +
  geom_boxplot(aes(x = treatment_skeleton, y = change,
                  fill = treatment_drupella),
               outlier.shape = NA) +
  scale_fill_manual(name = "Snail wound", labels = c("n" = "No", "y" = "Yes"),
                     values = c(no.snail.col, snail.col)) +
  geom_jitter(aes(x = treatment_skeleton, y = change,
                  color = treatment_drupella),
              # shape 21 with no fill: outline only, stroke set outside aes()
              size = 2, shape = 21, fill = NA, stroke = 0.5,
              position = position_jitterdodge(jitter.width = 0.15,
                                              jitter.height = 0,
                                              dodge.width = 0.5)) +
  scale_color_manual(name = "Snail wound", labels = c("n" = "No", "y" = "Yes"),
                     values = c(no.snail.col.dark, snail.col.dark)) +
  scale_x_discrete(labels = skel_labs) +
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        strip.background = element_blank(),
        legend.position = "top",
        axis.text.x = element_markdown(lineheight = 1)) +
  ylab("Weight change (g)") +
  xlab("") -> plt.a
```

Make panel b, left (snail marginal means)

```{r}
marg %>% 
  filter(factor == "Snail wound") %>% 
  ggplot() +
  geom_pointrange(aes(x = level_label, y = emmean,
                      ymin = lower.CL, ymax = upper.CL),
                  size = .8, linewidth = .8, color = "black") +
  geom_text(aes(x = level_label, y = upper.CL, label = group),
            vjust = -0.8, size = 4, color = "black") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.20))) +
  theme_classic() +
  theme(axis.line = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        axis.text.x = element_markdown(lineheight = 1)) +
  ylab("Weight change (g)") +
  xlab("Snail wound") -> plt.b
```

Make panel b, right (structure marginal means)

```{r}
marg %>% 
  filter(factor == "Structure treatment") %>% 
  ggplot() +
  geom_pointrange(aes(x = level_label, y = emmean,
                      ymin = lower.CL, ymax = upper.CL),
                  size = .8, linewidth = .8, color = "black") +
  geom_text(aes(x = level_label, y = upper.CL, label = group),
            vjust = -0.8, size = 4, color = "black") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.20))) +
  theme_classic() +
  theme(axis.line = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        axis.text.x = element_markdown(lineheight = 1),
        # drop the duplicated y-axis on the right-hand plot
        axis.title.y = element_blank()) +
  ylab(NULL) +
  xlab("Structure treatment") -> plt.c
```

Plot all panels together with raw data and save as a PDF in our `figures` folder

```{r}
pdf(here("figures", "fig_2_weights.pdf"), height = 5.5, width = 6)

# tags set directly so the two (b) sub-plots share one label
plt.a  <- plt.a  + labs(tag = "(a)")
plt.b <- plt.b + labs(tag = "(b)")


plt.a / (plt.b + plt.c + plot_layout(widths = c(2, 4))) +
  plot_layout(heights = c(1, 0.85))
dev.off()
```

