---
title: 'Analysis: Spring tissue change'
author: "Julianna Renzi"
date: "2026-03-26"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r clear environment}
# clear environment for reproducibility
rm(list=ls())
```

```{r load packages}
# load librarian (package for installing/loading packages)
if (!require("librarian")) install.packages("librarian")

# load required packages
librarian::shelf(here, # relative file paths 
                 tidyverse, # data wrangling
                 glmmTMB, # model creation
                 car, # type II ANOVA
                 emmeans, # marginal means
                 performance, # check model assumptions
                 multcomp, # compact letter displays
                 patchwork, # combining plots into one
                 scales # percent formatting
                 )
```

```{r set color palette}
"#E1BE6A" -> no.snail.col
"#40B0A6" -> snail.col

"#c49627" -> no.snail.col.dark
"#2d7d76" -> snail.col.dark
```

# Bring in data

```{r}
# tissue cover data
tissue_1 <- read_csv(here("data", "tissue_cover.csv"))

# metadata
metadata <- read_csv(here("data", "metadata.csv"))
```

# Prepare data

We sampled corals for microbes at the end of the 3-month experiment so we will use tissue cover data from after that sampling, on November 14, 2023 (`tissue_cover_postmicrobe`) compared with tissue cover data from the end of April/beginning of May (`tissue_cover_9mo`)

```{r}
tissue_1 %>% 
  # add metadata
  full_join(metadata) %>% 
  # filter out corals we need to exclude
  filter(is.na(exclude_date)) %>% 
  # don't consider corals that already lost 100% before this window
  filter(tissue_cover_postmicrobe != 0) %>% 
  # make proper skeleton treatments
  mutate(skeleton_yn = case_when(treatment_skeleton == "control" ~ "No",
                                 treatment_skeleton == "hali" ~ "Yes",
                                 treatment_skeleton == "loboph" ~ "Yes",
                                 treatment_skeleton == "skeleton" ~ "Yes")) %>% 
  # remove missing coral
  filter(!is.na(tissue_cover_9mo)) %>% 
  # make a column for full mortality
  mutate(total_mortality_yn = case_when(tissue_cover_9mo == 0 ~ 1,
                   tissue_cover_9mo > 0 ~ 0,
                   TRUE ~ -9999)) -> tissue_2
```

# Model mortality 

Model the probability of complete mortality (i.e., tissue cover = 0% after 9 months)

```{r}
m1 <- glmmTMB(total_mortality_yn ~ skeleton_yn + (1|parent_colony), 
              family = binomial,
              data = tissue_2)
```

Examine model fit

```{r}
check_model(m1)
```

Look at terms

```{r}
car::Anova(m1)
```

Examine the magnitude

```{r}
summary(m1) # coefficient for skeleton is 2.336

exp(2.336) # means the odds of “success” (i.e., dying) for a coral in the skeleton treatment is about 10.34 times higher than the odds of dying for a control coral
```

# Plot

## Panel a (data in bar plot)

Get sample sizes for top of dead bars

```{r}
# get sample sizes
tissue_2 %>% 
  mutate(mortality_binary = case_when(total_mortality_yn == 0 ~ "Alive", 
                                      total_mortality_yn == 1 ~ "Dead",
                                      TRUE ~ "ERROR")) %>% 
  # summarize
  group_by(skeleton_yn, mortality_binary) %>% 
  summarize(n = n()) %>% 
  pivot_wider(names_from = mortality_binary,
              values_from = n) %>% 
  mutate(proportion_corals = paste0(Dead, "/", Alive + Dead)) -> prop_corals
```

Plot

```{r}
tissue_2 %>% 
  mutate(mortality_binary = case_when(total_mortality_yn == 0 ~ "Alive", 
                                      total_mortality_yn == 1 ~ "Dead",
                                      TRUE ~ "ERROR")) %>% 
  ggplot(aes(x = skeleton_yn)) +
  scale_y_continuous(labels = percent) +
  geom_bar(stat = "count", position = "fill",
           aes(group = mortality_binary, fill = mortality_binary)) +
  ylab("Percent of corals") +
  xlab("") +
  scale_x_discrete(labels = c("No" = "Control", "Yes" = "Skeleton")) +
  scale_fill_manual(name = "Status in May", values = c("Alive" = "#e69f00", "Dead" = "#0072b2")) +
  theme_classic() +
  geom_text(data = prop_corals,
    aes(x = skeleton_yn, y = Dead/(Dead+Alive), label = proportion_corals),
    vjust = -0.5, color = "#0072b2"
  ) +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "top") -> mort.may.plt
```

## Panel b (marginal means)

```{r}
emm_skel  <- as_tibble(emmeans(m1, ~ skeleton_yn, type = "response")) # marginal means, averaged over snail
```

```{r}
emm_skel %>% 
  mutate(skeleton_yn = case_when(skeleton_yn == "No" ~ "Control",
                                 skeleton_yn == "Yes" ~ "Skeleton")) %>% 
  ggplot() +
  geom_pointrange(aes(x = skeleton_yn, y = prob,
                      ymin = asymp.LCL, ymax = asymp.UCL),
                  size = .8, linewidth = .8, color = "black") +
  theme_classic() +
  theme(axis.line = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        axis.text.x = element_markdown(lineheight = 1)) +
  ylab("Predicted probability of mortality") +
  xlab("") -> mort.emms.plt
```

## Plot together

```{r}
pdf(here("figures", "fig_4_mortality.pdf"), height = 3.5 , width = 7)

mort.may.plt + mort.emms.plt + plot_annotation(tag_levels = 'a',
                                            tag_prefix = '(',
                                            tag_suffix = ')')

dev.off()
```

# Confirm qualitatively the same if we keep all structure treatments in

```{r}
m2 <- glmmTMB(total_mortality_yn ~ treatment_skeleton + (1|parent_colony), 
              family = binomial,
              data = tissue_2)
```

```{r}
car::Anova(m2) 
```

Examine trends:

```{r}
plot(emmeans(m2, pairwise ~ treatment_skeleton))
```

See pattern is the same and the skeleton treatments all appear to behaving similarly, although some power is lost when we divide the skeleton treatments up

# Plot these effects for the supplement

Get marginal effects

```{r}
emm_skel <- emmeans(m2, ~ treatment_skeleton, type = "response") %>%  # marginal means, averaged over snail
  as_tibble() %>% 
  mutate(treatment_skeleton = case_when(treatment_skeleton == "control" ~ "Control",
                                        treatment_skeleton == "hali" ~ "<i>Halimeda</i><br>skeleton",
                                        treatment_skeleton == "loboph" ~ "<i>Lobophora</i><br>skeleton",
                                        treatment_skeleton == "skeleton" ~ "Skeleton"
                                        )) %>% 
  mutate(treatment_skeleton = factor(treatment_skeleton, 
                                     levels = c("Control", "Skeleton", "<i>Halimeda</i><br>skeleton", "<i>Lobophora</i><br>skeleton")))
```

Plot

```{r}
pdf(here("figures", "fig_S5_mortality_structure.pdf"), height = 3, width = 4)

emm_skel %>% 
  ggplot() +
  geom_pointrange(aes(x = treatment_skeleton, y = prob,
                      ymin = asymp.LCL, ymax = asymp.UCL),
                  size = .8, linewidth = .8, color = "black") +
  theme_classic() +
  theme(axis.line = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        axis.text.x = element_markdown(lineheight = 1)) +
  ylab("Predicted probability of mortality") +
  xlab("")

dev.off()
```

