---
title: 'Analysis: Wound healing'
author: "Julianna Renzi"
date: "2026-03-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r clear environment}
# clear environment for reproducibility
rm(list=ls())
```

```{r load packages}
# load librarian (package for installing/loading packages)
if (!require("librarian")) install.packages("librarian")

# load required packages
librarian::shelf(here, # relative file paths 
                 tidyverse, # data wrangling
                 performance, # check model assumptions
                 survival, # creating survival objects
                 survminer, # diagnostics
                 ggbreak, # broken axes
                 ggtext # axes lables
                 )
```

```{r set color palette}
"#E1BE6A" -> no.snail.col
"#40B0A6" -> snail.col
```

```{r}
# explicitly assign tidyverse functions to avoid confusion
summarize <- dplyr::summarize
group_by <- dplyr::group_by
select <- dplyr::select
```

# Read in data

Notes: 
- In the healing data 0 = not healed, 1 = healed
- *Drupella* were removed on **2023-07-21** 
- No wounds were healed on **2023-07-22** (first photographic time point)
- In the survival package a censored value of 1 = a true event and a value of 0 = a censored data point (e.g., study ended before event, removed from experiment)

```{r}
# wound healing data
healing_1 <- read_csv(here("data", "wound_healing.csv")) 

# wound size data
wound_size_1 <- read_csv(here("data", "wound_sizes.csv"))

# metadata
metadata <- read_csv(here("data", "metadata.csv"))
```

# Calculate time to event (healed wound)

First, join with metadata, get the dataframe into long format, and make date columns date objects

```{r}
healing_1 %>% 
  # make sure these are in order
  arrange(coral_id, observe_date) %>% 
  group_by(coral_id) %>% 
  # grab the last entry for each coral, which should be the last day observed, given that we used arrange(coral_id, date)
  slice_tail(n = 1) %>% 
  ungroup() %>% 
  # snails removed on 2023-07-21
  mutate(days_to_event = observe_date - as_date("2023-07-21")) %>% 
  mutate(days_numeric = as.numeric(days_to_event)) %>% 
  # join with wound size data 
  full_join(wound_size_1) %>% 
  # and metadata, but use inner_join() because only want snail treatments
  inner_join(metadata) -> healing_2
```

# Make Cox model

## With interaction

Make survival object

```{r}
surv.1 <- Surv(time = healing_2$days_numeric, # time to death
                    event = healing_2$censored) # Censored points: 0 = right censored (not a true death), 1 = event at time
```

Full model

```{r}
coxph.1 <- coxph(surv.1 ~ treatment_skeleton * size_mm2 + frailty(parent_colony), 
                    data = healing_2) # response here must be a survival object returned by the Surv funtion (why we make the surv_object above)
```

```{r}
summary(coxph.1)
```

## Get X^2's

Get X^2 using reduced model approach--can't use `car::Anova()` reliably here (want the likelihood ratio test not the Wald). First make reduced model:

```{r}
coxph.2 <- coxph(surv.1 ~ treatment_skeleton + size_mm2 + frailty(parent_colony), 
                    data = healing_2)
```

```{r}
anova(coxph.1, coxph.2) # no difference
```

So, can drop the interaction. Interpret the other coefficients:

```{r}
coxph.3 <- coxph(surv.1 ~ treatment_skeleton + frailty(parent_colony), 
                    data = healing_2)
```

```{r}
coxph.4 <- coxph(surv.1 ~ size_mm2 + frailty(parent_colony), 
                    data = healing_2)
```

Compare for **size** treatment (drop structure):

```{r}
anova(coxph.2, coxph.3) # important
```

And for **structure** treatment (drop size)

```{r}
anova(coxph.2, coxph.4) # not important
```

# Test model assumptions

We want to test that the hazards are proportional (don't change over time), which is assumed in the cox model. We'll use scaled Schoenfeld residuals, which are independent of time and should have a random pattern. The random assumption is supported by p > 0.05. 

```{r test Schoenfeld}
test.ph <- cox.zph(coxph.2) 
  test.ph # global meets assumption
  plot(test.ph)
```

Influential observations

```{r}
ggcoxdiagnostics(coxph.2, type = "dfbeta", # dfbeta plots the estimated changes in the regression coefficients upon deleting each observation in turn
                 ggtheme = theme_bw()) # looks like not much change
```

Nonlinearity assumption

```{r}
resid <- residuals(coxph.2, type="martingale") # for nonlinearity assumption, line should be horizontal and not angling

mr <- cbind(size_mm2 = healing_2$size_mm2, resid)
```

Plot--looks pretty horizontal and non-angling

```{r}
ggplot(data = mr,
       aes(x = size_mm2, y = resid)) +
  geom_point() + 
  geom_smooth() +
  xlab("Age") + 
  ylab("Martingale Residuals")
```

# Use time-varying hazards model

## Wrangle

Generally looks good, but size has a bit of a changing effect. Check to make sure results don't change with a time-varying hazards model

```{r}
healing_1 %>% 
  # calculate difference between the observed date and the start date
  mutate(tstart = case_when(observe_date == as_date("2023-07-22") ~ 0, 
                            TRUE ~ as.numeric(lag(observe_date) - as_date("2023-07-21")))) %>% 
  mutate(tstop = as.numeric(observe_date - as_date("2023-07-21"))) %>% 
  # and metadata, but use inner_join() because only want snail treatments
  inner_join(metadata) %>% 
  full_join(wound_size_1) -> healing_3
```

## Model

Full model:

```{r}
coxph.t.1 <- coxph(Surv(tstart, tstop, event = wound_healed) ~ treatment_skeleton + size_mm2 + size_mm2:treatment_skeleton + frailty(parent_colony), 
                   data = healing_3)
  summary(coxph.t.1)
```

Drop interaction:

```{r}
coxph.t.2 <- coxph(Surv(tstart, tstop, event = wound_healed) ~ treatment_skeleton + size_mm2 + frailty(parent_colony), 
                   data = healing_3)
  summary(coxph.t.2)
```

Drop each individual term

```{r}
coxph.t.3 <- coxph(Surv(tstart, tstop, event = wound_healed) ~ treatment_skeleton + frailty(parent_colony), 
                   data = healing_3)
```

```{r}
coxph.t.4 <- coxph(Surv(tstart, tstop, event = wound_healed) ~ size_mm2 + frailty(parent_colony), 
                   data = healing_3)
```

## Model comparisons

Compare interaction with no interaction

```{r}
anova(coxph.t.1, coxph.t.2) # no difference--drop interaction
```

Examine importance of size (i.e., dropped it--structure-only treatment):

```{r}
anova(coxph.t.1, coxph.t.3) # marginal
```

Examine importance of structure (i.e., dropped it--size-only treatment):

```{r}
anova(coxph.t.1, coxph.t.4) # not important
```

# Plot healing times

First, wrangle the dataframe for plotting

```{r}
# create axis value/"sentinel" for wounds that never healed
sentinel <- 200

# add this to the dataframe and classify points
healing_2 %>% 
  mutate(Status = case_when(censored == "1" ~ "Healed", 
                            (censored == "0" & days_numeric == 99) ~ "Never healed", 
                            TRUE ~ "Censored")) %>% 
  # get rid of corals that were removed
  filter(Status != "Censored") %>% 
  # create axis value/"sentinel" for wounds that never healed
  mutate(days_numeric = case_when(Status == "Never healed" ~ sentinel,
                                  TRUE ~ days_numeric)) -> healed_4
```

Plot

```{r}
pdf(here("figures", "fig_3_healing.pdf"), height = 4, width = 6.5, 
    onefile = FALSE) # remove white starter page

healed_4 %>% 
  ggplot(aes(x = size_mm2, y = days_numeric,
             color = treatment_skeleton)) +
  geom_point(size = 2, alpha = 0.7) +
  xlab(expression(paste("Wound size (", mm^2, ")"))) +
  ylab("Time to healing (days)") +
  scale_color_manual(name = "Structure treatment",
                      values = c("#440154", "#3B528B", "#35B779FF", "#FDE725"),
                      labels = (c("control" = "Control", "skeleton" = "Skeleton",
                "hali" = "<i>Halimeda</i> skeleton",
                "loboph" = "<i>Lobophora</i> skeleton"))
                      ) +
  # create y-axis break between real data max and the sentinel/unhealed values
  scale_y_break(breaks = c(100, 194), space = 0.2) +
  # replace the sentinel tick with text
  scale_y_continuous(
    breaks = c(0, 25, 50, 75, 100, sentinel),
    labels = c("0", "25", "50", "75", "100", "Never\nhealed"),
    limits = c(0, sentinel + 6)  # add breathing room above sentinel
  ) +
  scale_x_continuous(
    breaks = c(0, 50, 100, 150, 200, 250, 300)
  ) +
  theme_classic() +
  # get rid of ggbreaks second y-axis
  theme(axis.text.y.right  = element_blank(),
  axis.ticks.y.right = element_blank(),
  axis.line.y.right  = element_blank(),
  # move axes a little closer
  axis.title.y = element_text(margin = margin(t = 0, r = -14)),
  # and move x-axis label to be more centered on the actual plot
  axis.title.x = element_text(margin = margin(t = -4, r = 0), hjust = 0.37),
  panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "right", legend.text = element_markdown()) 

dev.off()
```





