---
title: 'Analysis: Spring tissue change'
author: "Julianna Renzi"
date: "2026-03-26"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r clear environment}
# clear environment for reproducibility
rm(list=ls())
```

```{r load packages}
# load librarian (package for installing/loading packages)
if (!require("librarian")) install.packages("librarian")

# load required packages
librarian::shelf(here, # relative file paths 
                 tidyverse, # data wrangling
                 glmmTMB, # model creation
                 car, # type II ANOVA
                 emmeans, # marginal means
                 performance, # check model assumptions
                 multcomp, # compact letter displays
                 patchwork, # combining plots into one
                 scales # percent formatting
                 )
```

```{r set color palette}
"#E1BE6A" -> no.snail.col
"#40B0A6" -> snail.col

"#c49627" -> no.snail.col.dark
"#2d7d76" -> snail.col.dark
```

# Bring in data

```{r}
# tissue cover data
tissue_1 <- read_csv(here("data", "tissue_cover.csv"))

# metadata
metadata <- read_csv(here("data", "metadata.csv"))
```

# Prepare data

We sampled corals for microbes at the end of the 3-month experiment so we will use tissue cover data from after that sampling, on November 14, 2023 (`tissue_cover_postmicrobe`) compared with tissue cover data from the end of April/beginning of May (`tissue_cover_9mo`)

```{r}
tissue_1 %>% 
  # add metadata
  full_join(metadata) %>% 
  # filter out corals we need to exclude
  filter(is.na(exclude_date)) %>% 
  # don't consider corals that already lost 100% before this window
  filter(tissue_cover_postmicrobe != 0) %>% 
  # make proper skeleton treatments
  mutate(skeleton_yn = case_when(treatment_skeleton == "control" ~ "No",
                                 treatment_skeleton == "hali" ~ "Yes",
                                 treatment_skeleton == "loboph" ~ "Yes",
                                 treatment_skeleton == "skeleton" ~ "Yes")) %>% 
  # calculate changes
  mutate(tissue_loss_pp = tissue_cover_postmicrobe - tissue_cover_9mo,
         tissue_loss_percent = (tissue_cover_postmicrobe - tissue_cover_9mo)/tissue_cover_postmicrobe) %>% 
  # remove missing coral
  filter(!is.na(tissue_cover_9mo)) %>% 
  # make a column for full mortality
  mutate(total_mortality_yn = case_when(tissue_cover_9mo == 0 ~ 1,
                   tissue_cover_9mo > 0 ~ 0,
                   TRUE ~ -9999)) -> tissue_2
```

# Model

## Tissue loss (PP change)

First, with the interaction

```{r}
m1 <- glmmTMB(tissue_loss_pp ~ skeleton_yn + treatment_drupella + skeleton_yn:treatment_drupella + (1|parent_colony), 
              family = gaussian,
              data = tissue_2)
```

```{r}
car::Anova(m1) # not important--drop it
```

Then, without the interaction

```{r}
m2 <- glmmTMB(tissue_loss_pp ~ skeleton_yn + treatment_drupella + (1|parent_colony), 
              family = gaussian,
              data = tissue_2)
```

```{r}
car::Anova(m2) 
```

Check magnitude

```{r}
emmeans(m2, pairwise ~ skeleton_yn)
```

Check model residuals

```{r}
check_model(m2) # not perfect, but should be OK
```

## Full mortality 

Also model the probability of complete mortality (i.e., tissue cover = 0% after 9 months)

```{r}
m3 <- glmmTMB(total_mortality_yn ~ skeleton_yn + treatment_drupella + skeleton_yn:treatment_drupella + (1|parent_colony), 
              family = binomial,
              data = tissue_2)
```

```{r}
car::Anova(m3)
```

Drop the interaction

```{r}
m4 <- glmmTMB(total_mortality_yn ~ skeleton_yn + treatment_drupella + (1|parent_colony), 
              family = binomial,
              data = tissue_2)
```

```{r}
car::Anova(m4)
```

Examine the magnitude

```{r}
summary(m4) # coefficient for skeleton is 2.418

exp(2.418) # means the odds of “success” (i.e., dying) for a coral in the skeleton treatment is about 11.22 times higher than the odds of dying for a control coral
```

# Plot

## Panel (a)

Get model dataframe

```{r}
grps <- as_tibble(cld(emmeans(m2, pairwise ~ skeleton_yn), Letters = letters)) %>% 
  # get rid of weird space
  mutate(group = str_remove_all(.group, " "))
```

```{r}
as_tibble(emmeans(m2, pairwise ~ skeleton_yn)$emmeans) %>% 
  full_join(grps) -> grps_2
```

Plot

```{r}
grps_2 %>% 
  ggplot() +
  geom_pointrange(aes(x = skeleton_yn, y = emmean, ymin = lower.CL, ymax = upper.CL),
                      size = .6, linewidth = .8) +
  # make a new color scale
  ggnewscale::new_scale_color() +
  geom_jitter(data = tissue_2,
              aes(x = skeleton_yn, y = tissue_loss_pp, color = treatment_drupella),
              width = 0.15, size = 2, shape = 1, height = 0) + # use hollow circles
  scale_color_manual(name = "Snail wound", 
                     labels = c("n" = "No", "y" = "Yes"),
                     values = c(no.snail.col.dark, snail.col.dark)) +
  theme_classic() +
  theme(panel.border = element_rect(color = "black", 
                                    fill = NA, linewidth = 1),
        legend.position = "top") +
  ylab("Tissue loss (pp change)") +
  xlab("") +
  scale_x_discrete(labels = c("none" = "Control", "skeleton" = "Skeleton")) +
  # make a new color scale
  ggnewscale::new_scale_color() +
  geom_text(data = grps_2,
            aes(label = group, x = skeleton_yn, y = upper.CL), 
              vjust = -1, size = 5, color = "black") -> pp.may.plt
```

## Panel (b)

Get sample sizes for top of dead bars

```{r}
# get sample sizes
tissue_2 %>% 
  mutate(mortality_binary = case_when(total_mortality_yn == 0 ~ "Alive", 
                                      total_mortality_yn == 1 ~ "Dead",
                                      TRUE ~ "ERROR")) %>% 
  # summarize
  group_by(skeleton_yn, mortality_binary) %>% 
  summarize(n = n()) %>% 
  pivot_wider(names_from = mortality_binary,
              values_from = n) %>% 
  mutate(proportion_corals = paste0(Dead, "/", Alive + Dead)) -> prop_corals
```

Plot

```{r}
tissue_2 %>% 
  mutate(mortality_binary = case_when(total_mortality_yn == 0 ~ "Alive", 
                                      total_mortality_yn == 1 ~ "Dead",
                                      TRUE ~ "ERROR")) %>% 
  ggplot(aes(x = skeleton_yn)) +
  scale_y_continuous(labels = percent) +
  geom_bar(stat = "count", position = "fill",
           aes(group = mortality_binary, fill = mortality_binary)) +
  ylab("Percent of corals") +
  xlab("") +
  scale_x_discrete(labels = c("No" = "Control", "Yes" = "Skeleton")) +
  scale_fill_manual(name = "Status in May", values = c("Alive" = "#e69f00", "Dead" = "#0072b2")) +
  theme_classic() +
  geom_text(data = prop_corals,
    aes(x = skeleton_yn, y = Dead/(Dead+Alive), label = proportion_corals),
    vjust = -0.5, color = "#0072b2"
  ) +
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
        legend.position = "top") -> mort.may.plt
```

### Alternative: model outputs

Get model dataframe

```{r}
grps1 <- as_tibble(cld(emmeans(m4, pairwise ~ skeleton_yn, type = "response"), Letters = letters)) %>%
  # get rid of weird space
  mutate(group = str_remove_all(.group, " "))
```

```{r}
as_tibble(emmeans(m4, pairwise ~ skeleton_yn, type = "response")$emmeans) %>% 
  full_join(grps) -> grps_3
```

Plot

```{r}
grps_3 %>% 
  ggplot() +
  geom_pointrange(aes(x = skeleton_yn, y = prob, ymin = asymp.LCL, ymax = asymp.UCL),
                      size = .6, linewidth = .8) +
  # make a new color scale
  ggnewscale::new_scale_color() +
  geom_jitter(data = tissue_2,
              aes(x = skeleton_yn, y = total_mortality_yn, color = treatment_drupella),
              width = 0.15, size = 2, shape = 1, height = 0) + # use hollow circles
  scale_color_manual(name = "Snail wound", 
                     labels = c("n" = "No", "y" = "Yes"),
                     values = c(no.snail.col.dark, snail.col.dark)) +
  theme_classic() +
  theme(panel.border = element_rect(color = "black", 
                                    fill = NA, linewidth = 1),
        legend.position = "top") +
  ylab("Probability of full mortality") +
  xlab("") +
  scale_x_discrete(labels = c("No" = "Control", "Yes" = "Skeleton")) +
  # make a new color scale
  ggnewscale::new_scale_color() +
  geom_text(data = grps_3,
            aes(label = group, x = skeleton_yn, y = asymp.UCL), 
              vjust = -1, size = 5, color = "black") -> mort.may.plt2
```

## Plot together

Option #1:

```{r}
pdf(here("figures", "fig_4_mortality.pdf"), height = 3.5 , width = 7)

pp.may.plt + mort.may.plt + plot_annotation(tag_levels = 'a',
                                            tag_prefix = '(',
                                            tag_suffix = ')')

dev.off()
```

Option #2:

```{r}
pdf(here("figures", "fig_4_mortality_2.pdf"), height = 3.5 , width = 7)

pp.may.plt + mort.may.plt2 + plot_annotation(tag_levels = 'a',
                                            tag_prefix = '(',
                                            tag_suffix = ')')

dev.off()
```

# Confirm qualitatively the same if other approaches used

## Percent change of remaining tissue

First, with the interaction

```{r}
m5 <- glmmTMB(tissue_loss_percent ~ skeleton_yn + treatment_drupella + skeleton_yn:treatment_drupella + (1|parent_colony), 
              family = gaussian,
              data = tissue_2)
```

```{r}
car::Anova(m5) # not important--drop it
```

Without the interaction

```{r}
m6 <- glmmTMB(tissue_loss_percent ~ skeleton_yn + treatment_drupella + (1|parent_colony), 
              family = gaussian,
              data = tissue_2)
```

```{r}
car::Anova(m6) # skeleton still important (even stronger this way)
```

## Absolute tissue cover in may 

First, with the interaction

```{r}
m7 <- glmmTMB(tissue_cover_9mo ~ skeleton_yn + treatment_drupella + skeleton_yn:treatment_drupella + (1|parent_colony), 
              family = gaussian,
              data = tissue_2)
```

```{r}
car::Anova(m7) # not important--drop it
```

Without the interaction

```{r}
m8 <- glmmTMB(tissue_cover_9mo ~ skeleton_yn + treatment_drupella + (1|parent_colony), 
              family = gaussian,
              data = tissue_2)
```

```{r}
car::Anova(m8) # skeleton still important (even stronger this way)
```

## Keep all structure treatments in

```{r}
m9 <- glmmTMB(tissue_loss_percent ~ treatment_skeleton + treatment_drupella + treatment_skeleton:treatment_drupella + (1|parent_colony), 
              family = gaussian,
              data = tissue_2)
```

```{r}
car::Anova(m9) # not important--drop it
```

Without the interaction

```{r}
m10 <- glmmTMB(tissue_loss_percent ~ treatment_skeleton + treatment_drupella + (1|parent_colony), 
              family = gaussian,
              data = tissue_2)
```

```{r}
car::Anova(m10) 
```

Examine trends:

```{r}
plot(emmeans(m10, pairwise ~ treatment_skeleton))
```

See pattern is the same and the skeleton treatments all appear to behaving similarly, although some power is lost

